---
title: "Phylogenetics in R"
subtitle: "Pol dataset"
author: "Louis du Plessis"
date: '`r format(Sys.time(), "Last modified: %d %b %Y")`'
output: html_document
params:
  polfile: "../../datasets/hiv/hiv_pol.fasta"
  
---

```{r setup, include=FALSE}

library(ape)
library(phangorn)

knitr::opts_chunk$set(echo = TRUE, fig.path="figs/", dev='png')


```


# Pol dataset

First we read in the alignment. 

```{r load-data}

pol <- read.dna(params$polfile, format = "fasta")

```

We can check to see that pol contains 42 sequences of length 689.
We can also display the names of the sequences contained in each.

We also extract the names of the sequences that stem from the patient, the victim, and the community.

```{r, eval=TRUE}
pol


labels(pol)

allpatients_pol <- sapply(strsplit(labels(pol), "\\|"), function(x) x[2])
patient_pol     <- labels(pol)[grep("^P", allpatients_pol)]
victim_pol      <- labels(pol)[grep("^V", allpatients_pol)]
community_pol   <- labels(pol)[grep("^LA", allpatients_pol)]

```

## Distance-based trees

Calculate a neighbour-joining tree using a TN93 substitution model. 

```{r pol-distancetrees}

D_pol_TN93 <- ape::dist.dna(pol, model="TN93", as.matrix=TRUE)
tree_pol_NJ_TN93 <- phangorn::NJ(D_pol_TN93)

```

Check if the tree is rooted and if not, plot the unrooted tree.

```{r pol-nj-unrooted}

if (!is.rooted(tree_pol_NJ_TN93)) {
    ape::plot.phylo(tree_pol_NJ_TN93, type = "unrooted", cex=0.5, lab4ut = "axial")
}

```

We assume the outgroup is `AY156771.1|LA02.RT` (one of the community sequences) and root the neighbour-joining tree on it, before plotting the rooted tree.

```{r root-nj}

  tree_pol_NJ_TN93_rooted <- root.phylo(tree_pol_NJ_TN93,  
                                        outgroup = "AY156771.1|LA02.RT ", 
                                        resolve.root = TRUE)
  plot.phylo(tree_pol_NJ_TN93_rooted, align.tip.label = TRUE, cex=0.5)
  
```

Now create a colour array to colour patient, victim and community sequences differently using the `tip.color` argument in `plot.phylo()`.

```{r plot-nj} 

    colours <- rep("black", length(allpatients_pol))
    colours[match(patient_pol, tree_pol_NJ_TN93_rooted$tip.label)] <- 'blue'
    colours[match(victim_pol, tree_pol_NJ_TN93_rooted$tip.label)] <- 'red'
    
    plot.phylo(tree_pol_NJ_TN93_rooted, align.tip.label = TRUE, cex=0.5, 
               tip.color = colours)
    tiplabels(NA, pch=16, col=colours)
    legend("bottomleft", legend=c("Patient", "Victim", "Environment"),
           col=c("blue", "red", "black"), pch=16, cex=0.8)

```

Build the neighbour-joining tree using a Jukes-Cantor model, plot the tree and 
then plot the tanglegram connecting the Jukes Cantor NJ tree to the TN93 NJ tree.

```{r tanglegram-nj}

# Construct and root Neighbour-joining tree
D_pol_JC <- ape::dist.dna(pol, model="JC", as.matrix=TRUE)
tree_pol_NJ_JC <- phangorn::NJ(D_pol_JC)
tree_pol_NJ_JC_rooted <- root.phylo(tree_pol_NJ_JC,  
                                    outgroup = "AY156771.1|LA02.RT ", 
                                    resolve.root = TRUE)

# Plot the tree
plot.phylo(tree_pol_NJ_JC_rooted, align.tip.label = TRUE, cex=0.5, 
           tip.color = colours)
legend("bottomleft", legend=c("Patient", "Victim", "Environment"),
           col=c("blue", "red", "black"), pch=16, cex=0.8)


# Plot the tanglegram
cophyloplot(tree_pol_NJ_JC_rooted, tree_pol_NJ_TN93_rooted, assoc=cbind(tree_pol_NJ_JC_rooted$tip.label, tree_pol_NJ_TN93_rooted$tip.label), show.tip.label = FALSE, space=100, gap =5, length.line=0.2)


```



## Parsimony

Calculate the parsimony score of the NJ tree. To use the function we have to first
convert the alignment to an object of type `phyDat`, which is used by the `phangorn` 
package. 

```{r parsimony-score}

  pol_phyDat <- as.phyDat(pol)  
  parsimony(tree_pol_NJ_JC, pol_phyDat)
```

Now use the `pratchet()` function to find the maximum parsimony tree and then use 
`acctran()` to adjust the branch lengths, and again plot the tree.

```{r maximum-parsimony}

  # Calculate the maximum parsimony tree
  tree_pol_MP <- pratchet(pol_phyDat)
  tree_pol_MP <- acctran(tree_pol_MP, data = pol_phyDat)
  
  # Plot it
  colours <- rep("black", length(tree_pol_MP$tip.label))
  colours[match(patient_pol, tree_pol_MP$tip.label)] <- 'blue'
  colours[match(victim_pol, tree_pol_MP$tip.label)] <- 'red'
    
  plot.phylo(tree_pol_MP, align.tip.label = TRUE, cex=0.7, 
               tip.color = colours)
  tiplabels(NA, pch=16, col=colours)
  legend("bottomleft", legend=c("Patient", "Victim", "Environment"),
       col=c("blue", "red", "black"), pch=16, cex=0.8)
```

We can also verify that the maximum parsimony tree has a lower parsimony score 
than the NJ tree. 

```{r parsimony-scores}

  parsimony(tree_pol_NJ_TN93, pol_phyDat)
  parsimony(tree_pol_MP, pol_phyDat)
  parsimony(tree_pol_NJ_TN93, pol_phyDat) < parsimony(tree_pol_MP, pol_phyDat)

```

Now we plot the tanglegram comparing the neighbour-joining tree to the maximum
parsimony tree.

```{r tanglegram-mp}

  # Root on the same sequence
  tree_pol_MP_rooted <- root.phylo(tree_pol_MP,  
                                   outgroup = "AY156771.1|LA02.RT ", 
                                   resolve.root = TRUE)
  
  # Plot the tanglegram
  cophyloplot(tree_pol_MP_rooted, tree_pol_NJ_TN93_rooted,
              assoc=cbind(tree_pol_MP_rooted$tip.label, 
                          tree_pol_NJ_TN93_rooted$tip.label), 
              show.tip.label = FALSE, space=100, gap =5, length.line=0.1)

```


## Maximum-likelihood

Calculate the maximum-likelihood tree using the GTR model. Since the tree is 
unrooted (check it!) also root it on the same outgroup. 

```{r ml-tree, eval=TRUE}

  tree_pol_ML_GTR <- pml_bb(pol, model="GTR")
  tree_pol_ML_GTR_rooted <- root.phylo(tree_pol_ML_GTR$tree,  
                                       outgroup = "AY156771.1|LA02.RT ", 
                                       resolve.root = TRUE)

```

Now plot the ML tree.

```{r plot-mltree}  

  colours <- rep("black", length(tree_pol_ML_GTR_rooted$tip.label))
  colours[match(patient_pol, tree_pol_ML_GTR_rooted$tip.label)] <- 'blue'
  colours[match(victim_pol, tree_pol_ML_GTR_rooted$tip.label)] <- 'red'
    
  plot.phylo(tree_pol_ML_GTR_rooted, align.tip.label = TRUE, cex=0.7, 
               tip.color = colours)
  tiplabels(NA, pch=16, col=colours)
  legend("bottomleft", legend=c("Patient", "Victim", "Environment"),
       col=c("blue", "red", "black"), pch=16, cex=0.8)

```


Tanglegram between the NJ and ML trees

```{r tanglegram-ml}

  # Plot the tanglegram
  cophyloplot(tree_pol_ML_GTR_rooted, tree_pol_NJ_TN93_rooted,
              assoc=cbind(tree_pol_ML_GTR_rooted$tip.label, 
                          tree_pol_NJ_TN93_rooted$tip.label), 
              show.tip.label = FALSE, space=100, gap =5, length.line=0.1)

```




